Mendelian randomization for causal inference in Genome-wide association studies

### 2022-02-23 mr-review.Rmd
added gmm text

Many approaches to MR can be classified into two categories: two-stage methods and 
likelihood-based methods. A two-stage statistical model involves two regression models [@burgess_review_2017;@sawa_exact_1969]. For a continuous 
outcome, one may perform "two-stage least squares", which involves two linear regressions. 
First, the exposure variable is regressed on the instrument (*i.e.*, the SNP genotypes) (Equation \@ref(eq:lm1)). We denote each subject's exposure variable value as $x_i$ in Equation \@ref(eq:lm1), while $\alpha_0$ denotes an intercept term and $\alpha_1$ is the slope. $z_i$ denotes the SNP instrument genotype for subject $i$, while $\epsilon_i$, the error term, is assumed to be independent among subjects and normally distributed with a shared common variance and mean zero.

@@

Likelihood-based MR methods, unlike two-stage methods, provide maximum likelihood 
estimates with their many desirable properties [@fisher1925theory]. Limited information maximum likelihood from econometrics 
is the earliest approach for likelihood-based inference in MR [@davidson1993estimation]. CoMM and 
PMR-Egger are two recent likelihood-based methods [@yang_comm_2019;@yuan_testing_2020]. We next discuss CoMM and CoMM-S2.

CoMM, along with its extension for summary statistics, CoMM-S2, uses a collaborative mixed model to 
relate exposure and outcome variables to the SNP genotypes [@yang_comm-s2_2020]. In this model, the 
authors jointly model two linear mixed effects models, one relating the outcome to the SNPs and a second 
tying the exposure to the SNPs. Model fitting then 
proceeds via a parameter-expanded EM algorithm [@yang_comm_2019;@liu_parameter_1998]. @yang_comm_2019, 
working with gene expression traits as exposures, found that CoMM better prioritized genetic variants 
relative to other methods [@yang_comm_2019].

>>>>

Approaches to MR can be classified into four categories: ratio of coefficients method, two-stage methods, 
likelihood-based methods, and semiparametric methods.

### Ratio of coefficients method

The ratio of coefficients method, also known as the Wald method, estimates the 
causal effect of the exposure X on the outcome Y by using a single SNP instrument [@wald_fitting_1940]. For a continuous outcome, the causal effect estimator, $\hat\gamma_{Wald}$ is defined in Equation \@ref(eq:wald).

\begin{equation}
  \hat\gamma_{Wald} = \frac{\hat\beta_{YZ}}{\hat\beta_{XZ}}
  (\#eq:wald)
\end{equation}

In Equation \@ref(eq:wald), $\hat \beta_{YZ}$ and $\hat \beta_{XZ}$ are the slope 
estimates from the regressions of the outcome and exposure, respectively, on the 
SNP instrument. Since the Wald method requires only the regression coefficients, it can be used with summary data. However, the Wald method doesn't accommodate multiple SNP instruments, which limits its use in the GWAS setting.

### Two-stage methods

A two-stage statistical model involves two regression models [@burgess_review_2017;@sawa_exact_1969]. For a continuous 
outcome, one may perform "two-stage least squares", which involves two linear regressions. 
First, the exposure variable is regressed on the instrument (*i.e.*, the SNP genotypes) (Equation \@ref(eq:lm1)). We denote each subject's exposure variable value as $x_i$ in Equation \@ref(eq:lm1), while $\alpha_0$ denotes an intercept term and $\alpha_1$ is the slope. $z_i$ denotes the SNP instrument genotype for subject $i$, while $\epsilon_i$, the error term, is assumed to be independent among subjects and normally distributed with a shared common variance and mean zero.

@@

### Likelihood-based methods

Likelihood-based MR methods, unlike two-stage methods, provide maximum likelihood 
estimates with their many desirable properties [@fisher1925theory]. Limited information maximum likelihood from econometrics 
is the earliest approach for likelihood-based inference in MR [@davidson1993estimation].

Limited information maximum likelihood with a single SNP instrument is modeled with two equations (Equations \@ref(eq:liml1) and \@ref(eq:liml2)), where the random errors follow a bivariate normal distribution [@burgess_review_2017].

\begin{equation}
  x_i = \alpha_0 + \alpha_1 z_{i1} + \epsilon_{Xi}
  (\#eq:liml1)
\end{equation}

\begin{equation}
  y_i = \beta_0 + \beta_1 x_i + \epsilon_{Yi}
  (\#eq:liml2)
\end{equation}

Limited information maximum likelihood is sometimes called the maximum likelihood 
counterpart of two-stage least squares, and it yields the same causal estimate as 
two-stage least squares and the ratio method when used with a single SNP 
instrument. Additionally, the limited information maximum likelihood framework can accommodate $K$ SNP instruments by replacing $\alpha_1 z_{i1}$ with the sum $\sum_{k = 1}^K\alpha_k z_{ik}$.

One may also use Bayesian methods to obtain likelihood-based estimators [@kleibergen_bayesian_2003]. @burgess_improving_2012 examined a model that is similar to that from the limited information maximum likelihood framework (Equation \@ref(eq:b1)). The Bayesian model differs from the limited information maximum likelihood model in that the causal effect parameter, $\beta_1$ represents the effect between the true means for the exposure and the outcome. In the limited information maximum likelihood model, the causal effect is that of the measured effect on the measured outcome.

\begin{eqnarray}
\begin{pmatrix}
X_i \\
Y_i
\end{pmatrix} \sim N_2\left(\begin{pmatrix} \xi_i\\
\eta_i
\end{pmatrix}, \Sigma
\right)\nonumber \\
\xi_i = \alpha_0 + \alpha_1 z_{i1}\nonumber \\
\eta_i = \beta_0 + \beta_1 \xi_i
  (\#eq:b1)
\end{eqnarray}

For each subject $i$, the exposure and outcome values come from a bivariate normal 
distribution. They assumed that the mean of the exposure distribution is linear in 
the SNP instrument and that the mean of the outcome distribution is a linear in the 
mean exposure [@burgess_review_2017;@jones_choice_2012]. 
@burgess_improving_2012 demonstrated that the Bayesian model, with weak 
instruments, outperforms the limited information maximum likelihood model in terms 
of frequentist coverage levels.

### Semiparametric methods

Semiparametric instrumental variable methods, which feature parametric and nonparametric components, 
typically assume a parametric model connecting the exposure and outcome, but don't 
impose distributional assumptions on the errors [@burgess_review_2017]. Compared to 
fully parametric models, semiparametric models often are more robust to model 
misspecification [@burgess_review_2017;@clarke_instrumental_2012]. We follow 
@burgess_review_2017 by discussing three semiparametric strategies, generalized 
method of moments, continuous updating estimator, and G-estimation of structural 
mean models.

Generalized method of moments can be viewed as a more flexible form of two-stage 
least squares that handles heteroscedastic errors and nonlinearity in the two 
regressions [@burgess_review_2017]. Before specifying the model, we introduce 
notation. $\mathbb{E}(Y|\text{do}(X = x))$ is the conditional expectation of $Y$ if 
we forced $X$ to take value $x$ for every subject [@pearl2009causality].

The generalized method of moments equations, with a single SNP instrument, then, can be written as in Equation \@ref(eq:gmm).

\begin{eqnarray}
  \sum_i (y_i - f(y_i;\beta)) = 0\nonumber \\
  \sum_i z_i(y_i - f(y_i;\beta)) = 0
  (\#eq:gmm)
\end{eqnarray}

The GMM estimate is the value of the vector $\beta$ that sastisfies Equation \@ref(eq:gmm), where $f(\mathbb{E}(Y|\text{do}(X = x))$
